---
title: "Assignment--1"
author: "Hooman Fareghi"
date: "2026-01-25"
output: html_document
---
#Data source (downloaded from): https://practicedatasets.online/sales-dataset-for-analysis Selected Dataset Type: Customer Information #Number of Records: 5000

#Import data


```{r}
library(readxl)
data<- read_excel("customer_info (1).xlsx")
```

#Descriptive statistics for age.

```{r}
summary(data$Age)
```
#Descriptive statistics for Total spend.
```{r}
summary(data$`Total Spend ($)`)
```
#Descriptive statistics for average purchase amount.
```{r}
summary(data$`Average Purchase Amount ($)`)
```



#Descriptive statistics for gender.

```{r}
table(data$Gender)
```
#Descriptive statistics for region.
```{r}
table(data$Region)
```

#Descriptive statistics for preferred payment method.
```{r}
table(data$`Preferred Payment Method`)
```

#Transformmed age to age group.

```{r}
data$age_group <- cut(data$Age, breaks = c(-Inf , 29, 55, Inf) , labels = c("young(under30)" , "adult(30-55)", "senior(+55)"))
table(data$age_group)
```

#Transformmed total spend to spending level.
```{r}
data$spending_level <- cut(data$`Total Spend ($)`, breaks = c(-Inf, 2000, 10000, Inf), labels= c("low","medium","high"))
table(data$spending_level)
```



##plot of age (small breaks).


```{r}
hist(data$Age, main = "plot of age", xlab = "age" , col = "blue")
```



#Plot of purchase history (small breaks)


```{r}
hist(data$`Purchase History (count)`, main = "plot of purchase history", xlab = "purchase history" , col = "purple")
```



#Plot of purchase history (large breaks)


```{r}
hist(data$`Purchase History (count)`, breaks = 24 , main = "plot of purchase history", xlab = "purchase history" , col = "gray")
```


#Plot of age (large breaks)


```{r}
hist(data$Age, breaks = 24 , main = "plot of age", xlab = "age" , col = "lightgreen")
```


#Scatterplot of age vs total spend.


```{r}
plot(data$Age, data$`Total Spend ($)`, main = "age vs total spending", xlab ="age", ylab = "total spending" , col= "black")
```



#Scaterplot  purchase history vs total spend 



```{r}
plot(data$`Purchase History (count)`, data$`Total Spend ($)`,  main = "purchase history(count) vs total spending", xlab ="purchase history (count)", ylab = "total spending" , col= "darkgreen")
```


#Scaterplot age vs average purchase amount


```{r}
plot(data$Age, data$`Average Purchase Amount ($)`, main = "age vs Average Purchase Amount ($)", xlab ="age", ylab = "Average Purchase Amount ($)" , col= "lightblue")
