---
title: "Assignment4"
author: "Hooman Fareghi"
date: "2026-03-26"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#load and choose numerical variables and standardize them
```{r}
library(dplyr)
library(ggplot2)
library(cluster)
pharma <- read.csv("Pharmaceuticals-1.csv")
numV <- pharma %>%select(Market_Cap, Beta, PE_Ratio, ROE, ROA,Asset_Turnover, Leverage, Rev_Growth, Net_Profit_Margin)
numVS <- numV %>%scale()
```


#(A)use elbow method
```{r}
wss <- 1:10 %>%sapply(function(k) {kmeans(numVS, centers = k, nstart = 25)$tot.withinss})
data.frame(k = 1:10, wss = wss) %>%ggplot(aes(x = k, y = wss)) +geom_line() +geom_point() +
labs(title = "Elbow Method",x = "Number of Clusters",y = "Within-Cluster Sum of Squares")
```


#use silhouette method
```{r}
avg_sil <- 2:10 %>%sapply(function(k) {km <- numVS %>%kmeans(centers = k, nstart = 25)
    silhouette(km$cluster, dist(numVS))[, 3] %>% mean()})
data.frame(k = 2:10, silhouette = avg_sil) %>%ggplot(aes(x = k, y = silhouette)) +geom_line() +
  geom_point() +labs(title = "Silhouette Method",x = "Number of Clusters",y = "Average Silhouette Width")
```

#use k=3 and add cluster to data and show which firm is in which cluster
```{r}
set.seed(123)
km_final <- numVS %>%kmeans(centers = 3, nstart = 25)
pharma <- pharma %>%mutate(Cluster = as.factor(km_final$cluster))
pharma %>%select(Symbol, Name, Cluster)
```

#(B)Interpret the clusters using the numerical variables

```{r}
cluster2 <- pharma %>%group_by(Cluster) %>%summarise(across(c(Market_Cap, Beta, PE_Ratio, ROE, ROA,
  Asset_Turnover, Leverage, Rev_Growth, Net_Profit_Margin),mean, na.rm = TRUE))
cluster2
```


#(C) Check pattern for variables not used in clustering

```{r}
pharma %>%count(Cluster, Median_Recommendation)
pharma %>%count(Cluster, Location)
pharma %>%count(Cluster, Exchange)
```
#(D) Provide an appropriate name for each cluster using any or all of the variables in the dataset.
```{r}
clusterN <- cluster2 %>%mutate(ClusterN= case_when(Market_Cap == max(Market_Cap) ~ "Large Firms",Rev_Growth == max(Rev_Growth) ~ "Growth Firms",TRUE ~ "Other Firms"))
pharma <- pharma %>%left_join(clusterN%>% select(Cluster, ClusterN), by = "Cluster")
clusterN %>%select(Cluster, ClusterN)
```




